---
title: 'INFX 573: Problem Set 1 - Exploring Data'
author: TAPASVI BANSAL
date: 'Due: Thursday, October 12, 2017'
output: pdf_document
---

<!-- This syntax can be used to add comments that are ignored during knitting process. -->

##### Collaborators: Ishaan Srivastava

##### Instructions: #####

Before beginning this assignment, please ensure you have access to R and RStudio. 

1. Download the `problemset1.Rmd` file from Canvas. Open `problemset1.Rmd` in RStudio and supply your solutions to the assignment by editing `problemset1.Rmd`. 

2. Replace the "Insert Your Name Here" text in the `author:` field with your own full name. Any collaborators must be listed on the top of your assignment. 

3. Be sure to include well-documented (e.g. commented) code chucks, figures and clearly written text chunk explanations as necessary. Any figures should be clearly labeled and appropriately referenced within the text. 

4. Collaboration on problem sets is acceptable, and even encouraged, but each student must turn in an individual write-up in his or her own words and his or her own work. The names of all collaborators must be listed on each assignment. Do not copy-and-paste from other students' responses or code.

5. When you have completed the assignment and have **checked** that your code both runs in the Console and knits correctly when you click `Knit PDF`, rename the R Markdown file to `YourLastName_YourFirstName_ps1.Rmd`, knit a PDF and submit the PDF file on Canvas.

*stress more visualization, dplyr, less questions/ethics, etc*

##### Setup: #####

In this problem set you will need, at minimum, the following R packages.

```{r Setup, message=FALSE}
# Load standard libraries
library("tidyverse")
library("nycflights13")
```

#### Problem 1: Exploring the NYC Flights Data ####

In this problem set we will use the data on all flights that departed NYC (i.e. JFK, LGA or EWR) in 2013. You can find this data in the `nycflights13` R package. 

##### (a) Importing and Inspecting Data: ##### 
Load the data and describe in a short paragraph how the data was collected and what each variable represents. Perform a basic inspection of the data and discuss what you find.


```{r}
library(nycflights13)
library (magrittr)
library (dplyr)
library(tidyverse)
?nycflights13::flights
?nycflights13::airports
destsea2 <- structure(list())
head(nycflights13::flights)
View(nycflights13::planes)
table(nycflights13::planes$speed)
View(nycflights13::weather)
table(nycflights13::weather$origin)
nycflights13::flights %>% filter(dest == "SEA" & month == 12)
```

The package data describes the on-time flights data that departed out of NYC airports. The data has 5 individual tabular data with individual detailed focus. The airlines table has carrier code and assoicated airlines name. The planes table talks about the details about planes like manufacturer, type, engine, etc. With majority of planes belonging to Airbus (336), Airbus Industries (400),Bombardier (368), Embraer(299) and majority of "speed" data is missing.
The airports table details on location, FAA airport code, etc. The weather data talks about the weather in relation to the three origin airports (EWR, JFK, LGA). The flights data provides in depth information about the various flights that took off at the three airports. 


##### (b) Formulating Questions: ##### 
Consider the NYC flights data. Formulate three motivating questions you want to explore using this data and explain why they are of interest. 

Question 1:
a)Numbers are always interesting, the most elementary information that motivates me for this partiuclar data is what is the frequency of flights to Seattle, flying out of NYC in different months?
b)Another close question that fairly affects a travelers decision is the flight frequency by various carriers i.e what is the frequency of flights to Seattle offered by various existing air carriers?


Question 2:
When taking flights, it could be fruitful to know the limitations and general numbers on arrival delays exhibited by flights belonging to various air carriers i.e What is the pattern/distribution of arrival delay in Seattle exhibited by various airline carriers?


Question 3:
In general we look into the information that appears to directly affect us, the airlines organizations serve merely as a channel between passesngers and airplanes. It would be interesting to know about What are the current and past manufacturers of airplanes, the nature of planes they produced and the years they have been operating in?



##### (c) Exploring Data: ##### 

For each of the questions you proposed in Problem 1b, perform an exploratory data analysis designed to address the question. At a minimum, you should produce two visualizations related to each question. Be sure to describe what the visuals show and how they speak to your question of interest. 

Question 1:

```{r}
head(nycflights13::flights)
nycflights13::flights %>% filter(dest == "SEA" & month == 8)
nycflights13::flights %>% filter(dest == "SEA" & carrier== "AA")
destsea <- na.omit (nycflights13::flights %>% filter(dest == "SEA"))

x <- data.frame(table(destsea$month))
y <- data.frame(table(destsea$carrier))

ggplot(data= x %>% mutate (month = Var1, No._flight=Freq)) + geom_point(aes(x= month, y= No._flight), col = "purple")   #Plotting graph between the frequency of flights and the 12 months


ggplot(data= y %>% mutate (Carrier = Var1, No._flight=Freq)) + geom_point(aes(x= Carrier, y= No._flight),col = "red")  # Plotting graph between the frequency of flights and the different relevant carriers

```

a)Through the analysis of the plot above that is based on the data provided, we witness that the frequency of flights to Seattle, flying out of NYC is maximum in the month of July (431) and it is lowest in the second month of the year i.e February (224).
b)Out of the 5 major carriers, we observe that Delta Air Lines Inc. offers the highest frequency of flights to Seattle from NYC (1213) whereas American Airlines Inc. offers the least number of flights to Seattle from NYC (365).

##################################################################################################


Question 2:

```{r}

ggplot(data= destsea %>% mutate (minutes=arr_delay)) + geom_boxplot(aes(x= carrier, y= minutes), col= "red")
# Plotting graph between the arrival delays and various carriers for the flights from NYC to Seattle.

boxplot(destsea$arr_delay~destsea$carrier, ylim = c(-30,30), ylab = "Minutes", col="blue")

# Using boxplot to Plot graph between the arrival delays and various carriers for the flights from NYC to Seattle.


```
In order to observe the pattern/distribution of arrival delay in Seattle exhibited by various airline carriers, I used box-whisker plot to observe the variability outside the upper and lower quartiles
in addition to observe the central tendency of the data. JetBlue Airways (B6) appears to display the highest of all delays. Also, the relatively higher median projects that Jetblue is more consistent at arrival delays from NYC to Seattle.
With the upper quartile for arrival delay data falling on 0 and the Maximum of box-whisker plot falling relatively below of all the other airlines involved, Delta Air Lines Inc. seems to be more consistent at arriving on time in Seattle from NYC. Delta also seems to have more spreaded out outliers above the maximum and reflects outliers below the minimum of box-whisker plot.


##################################################################################################

Question 3:

```{r}

planesdata <- nycflights13::planes
ggplot(data= planes ) + geom_point(aes(x= year, y= manufacturer, col= engine))

```
It seems interesting to know about manufacturers in this data, Embraer, Bombardier Inc, Boeing, Airbus appears to be the major and continuous manufacturers of Turbo-fan type engine planes recently.
Like Boeing that has been continuously operating over two decades and has been majorly producing Turbo-fan and Turbo-jet type of engine based airplanes.
The manufacturer Beech seems to be the only Turbo-Prop type engine planes manufactuer whose production existed mostly between late 1960s to early 1970s.



##### (d) Challenge Your Results: ##### 

After completing the exploratory analysis from Problem 1c, do you have any concerns about your findings?

Due to limitations of time, I missed exploring about the various factors that are related to each other like the weather patterns and/or time of day associated with the delays in departure and arrival of flights.
Due to limitation of current skillset, I missed connecting dots between the location of airports and/or weather to arrival or departure delays.

Some of my feeble attempts of making a sense out of departure delays and various carriers or departure delays and flight origin.

```{r}
nycflights<-na.omit(nycflights13::flights)

ggplot(data= nycflights %>% mutate (minutes=dep_delay)) + geom_boxplot(aes(x= origin, y= minutes), col= "red")

boxplot(nycflights$dep_delay~nycflights$carrier, ylim = c(-30,30) ,ylab = "Minutes")

boxplot(nycflights$dep_delay~nycflights$origin, ylim = c(min(nycflights$dep_delay),max (nycflights$dep_delay)) ,ylab = "Minutes")

```

